---
title: "data wrangling intro"
author: "Joe DeCesaro"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(tidyverse)
```

### Warm-up

look at 'us-rent-income' 
```{r}
us_rent_tidy <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable, values_from = estimate)
```

### Practice some tidyverse functions for wrangling data
- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize`

### dplyr::filter() to keep of exclude rows based on our conditions

1. exact match single condition: only keep observations where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```

2. "and" statement within filter

Only keep observations if the penguin species is Chinstrap AND the year is 2008
```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == "2008")
# comma does the exact same thing as an "&" in the filter() function
```

3. Exclude Gentoos

Keep observations if the penguin species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Exercise:
Create a subset from "penguins" that only contains observations for Adelie penguins with a flipper length greater than or equal to 185mm.

```{r}
excercise_1 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)
```

4. the %in% operator

Use the %in% operator to look for *any* operator that matches within a vector, and return TRUE if the value does not exist for our row.

We want to keep observations where the island is Dream or Torgersen
```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
# If you used the == with a vector it would look for those two values in order of each other. We do not want to do that.
```

### select() to select or exclude COLUMNS

5. Keep a single column by name 
Only want to keep the species column.
```{r}
ex_5 <- penguins %>% 
  select(species)
# reqrite without the pipe operator

ex_5.1 <- select(.data = penguins, species)
```

6. Keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year
```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```


7. Choose or exclude columns based on position
```{r}
ex_7 <- penguins %>% 
  select(-4)
```


8. A piped sequence

Write a piped sequence where you first only keep columns for species, body_mass_g, and island second: filter to only include observations if body_mass_g is greater than 3600
```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```


### mutate() to add or update existing columns, while keeping the ones already there

9. add a column that converts body mass to kg
```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm/ 1000)
```

10. mutate to update an existing column
```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HORRAY!"))
```


11. Update the class of the bill_depth column into character
```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_11$bill_depth_mm)
```

### case_when(): your favorite if_else function

12. Create a new column that contains "small" if a penguin flipper length is less than 185, "medium" otherwise
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small", 
    flipper_length_mm >= 185 ~ "medium"
  ))
```

Task: starting from penguins, create a new column, bird_bill_bin that contains:

-If bill length < 38mm: "short bill"
-If bill length >= 38mm & <43mm: "medium bill"
-Id bill length >= 43 : "long bill"
```{r}
task_1 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
  ))
```

### group_by() %>% summarize() to find summary values by group, and return them in a nice summary table

13. group_by() to say there are groups within the species column
```{r}
ex_13 <-penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))
```

### Practice
in a single piped sequence:

- Only keep penguins that have a body mass > 3500 g
- Only keep columns for species, flipper_length_mm, and sex
- Create a summary table that contains the mean, sd, max and min value of flipper length by species and sex (as grouping variables)

```{r}
practice_1 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flipper = max(flipper_length_mm, na.rm = TRUE),
            min_flipper = min(flipper_length_mm, na.rm = TRUE))
```

